Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system of plausible mechanism

### 2024-08-13 _tex/index.tex
Built site for gh-pages

According to \citet{Verhavert_et_al_2019} and \citet{vanDaal_2020},
several assessment factors interact and influence the method's outcome.
These factors include the number and characteristics of the stimuli,
their \emph{proximity} in terms of the assessed trait, the number of
comparison per stimulus, and the pairing algorithm used. Furthermore,
since the method relies on judges' assessments, the number and
characteristics of judges, their \emph{discrimination} abilities, and
the number of comparisons per judge also play pivotal roles. Moreover,
when the stimuli represent sub-units of higher-levels units, factors
such as the number and characteristics of these units, along with their
\emph{proximity} in terms of the assessed trait, can significantly
influence the outcome. For instance, \citet{vanDaal_et_al_2016} assessed
academic writing skills of university students (units) using multiple
argumentative essays (sub-units).

Although several studies have examined the individual impact of these
factors on the method's reliability
\citep{Bramley_2015, Pollitt_2012b, Bramley_et_al_2019, Verhavert_et_al_2019, Crompvoets_et_al_2022, vanDaal_et_al_2017, Gijsen_et_al_2021, Bouwer_et_al_2023},
none, to the best of the authors' knowledge, have provided a transparent
depiction of the DCJ system and the mechanisms generating the DCJ
outcome. This study aims to fill this gap by representing DCJ within the
framework of causal inference. Specifically, utilizing the structural
approach, the study develops a scientific model to clarify plausible
causal assumptions and mechanisms inherent in the DCJ system. The study
then translates the scientific model into a probabilistic statistical
model. This model aims to produce statistical estimates to draw
inferences about plausible causal relationships within the DCJ system.

@@

This section introduces fundamental concepts in causal inference, such
as directed acyclic graphs (DAGs), structural causal models (SCMs), and
the flow of association and causation in graphs. The section is not a
comprehensive description of causal inference methods. Readers
interested in deeper exploration should consult introductory papers like
\citet{Pearl_2010}, \citet{Rohrer_2018}, \citet{Pearl_2019}, and
\citet{Cinelli_et_al_2020}. They may also find introductory books such
as \citet{Pearl_et_al_2018}, \citet{Neal_2020} and
\citet{McElreath_2020} useful. For more advanced study, seminal
intermediate papers like \citet{Neyman_et_al_1923}, \citet{Rubin_1974},
\citet{Spirtes_et_al_1991}, and \citet{Sekhon_2009}, as well as books
such as \citet{Pearl_2009}, \citet{Morgan_et_al_2014} and
\citet{Hernan_et_al_2020} are recommended.

\subsection{The empirical research
endeavor}\label{sec-framework-empirical}

The goal of an investigator is to generate new knowledge or deepen the
understanding of specific topics or issues. Investigators accomplish
this by addressing research problems. Research problems fall into two
categories: empirical and non-empirical research. The first tackles
real-world challenges, relying on evidence gathered through observation
and experimentation. The latter focuses on conceptual inquiries that
enhance understanding complex concepts, theories, and phenomena.

Addressing research problems requires formulating well-defined research
questions that provide focus, clarity, and structure to a study. In
empirical research, these questions are often framed as \emph{targets of
inference} or \emph{estimands}, representing the quantities the
investigator aims to estimate. For instance, an investigator might want
to examine how the duration of cochlear implant use affects the average
overall speech quality of a child.

Furthermore, in empirical research, investigators must design
appropriate methodologies, collect data, and ensure a systematic and
coherent analysis of the results. To achieve this, they must accurately
identify the factors influencing an outcome and correctly attribute the
effects to each factor. Consequently, understanding how an investigator
transitions from a target of inference to generating statistical
estimates is crucial.

Figure~\ref{fig-IEflow} shows the Identification-Estimation flowchart
\citep{McElreath_2020, Neal_2020}, which illustrates the mechanism for
transitioning from a estimand to a corresponding \emph{statistical
estimate}, through identification and estimation.

@@

\caption{\label{fig-IEflow}Identification-Estimation flowchart.
Extracted from \citet[32]{Neal_2020}}

@@

In statistics, \emph{causal inference} refers to the process of
identifying the causes of a phenomenon and estimating their effects
using data \citep{Shaughnessy_et_al_2010, Neal_2020}. Unlike classical
statistical modeling, which focuses solely on summarizing data and
inferring associations, causal inference provides a coherent
mathematical notation for analyzing causes and counterfactuals
\citep{Pearl_2009}.

Counterfactuals represent scenarios \emph{contrary to fact}, where
alternative \emph{potential} outcomes resulting from a cause are neither
observed nor observable \citep{Neal_2020, Counterfactual_2024}.
According to \citet{Pearl_et_al_2018}, counterfactuals are the
foundation of causal inference and occupy the highest level of cognitive
abstraction in the ladder of causation, followed by intervention and
association. Nevertheless, despite their abstract nature,
counterfactuals enable the development of a \emph{theory of the world}
that explains why specific causes have specific effects and what occurs
in their absence \citep{Pearl_et_al_2018}. They achieve this by
translating causal statements into counterfactual statements, that is,
statements about ``what would have happened in the world under different
circumstances.''

Several approaches to causal inference and counterfactuals exist, but
two are particularly prominent: the potential outcomes approach, also
known as the Neyman-Rubin causal model
\citep{Neyman_et_al_1923, Rubin_1974}, and the structural approach
\citep{Pearl_2009, Pearl_et_al_2016}. Both approaches employ rigorous
mathematical notation to characterize causal inference, but they do so
in different ways \citep{Neal_2020}. The potential outcomes approach
relies on counterfactual notation, whereas the structural approach
employs the do-operator and structural causal models
\citep[SCM,][]{Pearl_2009, Pearl_et_al_2016}. Despite these differences,
both notations can be expressed in terms of the other, and both
approaches provide methods for using experimental and observational data
to estimate causal effects \citep{Pearl_2010}.

However, the structural approach offers an additional key advantage over
the potential outcomes approach: it enables the graphical representation
of any system through directed acyclic graphs
\citep[DAG,][]{Gross_et_al_2018, Neal_2020}. DAGs serve as heuristics,
effectively conveying the assumed causal structure of a system
\citep{Pearl_et_al_2016}. A heuristic is a strategy that simplifies
information to make decisions more quickly, efficiently, and sometimes
more accurately than complex methods \citep{Chow_2015}. Consequently,
DAGs do not represent detailed statistical models but allow researchers
to deduce which statistical models can provide valid causal inferences,
assuming the causal structure depicted in the DAGs are accurate
\citep{McElreath_2020}.

@@

\includegraphics[width=0.9\textwidth,height=\textheight]{images/figures/ACflow.png}

@@

\caption{\label{fig-ACflow}The flow of association and causation in
graphs. Extracted from \citet[31]{Neal_2020}}

>>>>

According to \citet{Verhavert_et_al_2019} and \citet{vanDaal_2020},
several assessment factors interact and influence the method's outcome.
These factors include the number and characteristics of the stimuli,
their \emph{proximity} in terms of the assessed trait, the number of
comparison per stimulus, and the pairing algorithm used. Furthermore,
since the method relies on judges' assessments, the number and
characteristics of judges, their \emph{discrimination} abilities, and
the number of comparisons per judge also play pivotal roles. Moreover,
when the stimuli represent sub-units of higher-levels units, factors
such as the number and characteristics of these units, along with their
\emph{proximity} in terms of the assessed trait, can significantly
influence the outcome. For example, in \citet{vanDaal_et_al_2016}, the
authors assessed the academic writing skills of university students
(units) using multiple argumentative essays (sub-units).

Although several studies have examined the individual impact of these
factors on the method's reliability
\citep{Bramley_2015, Pollitt_2012b, Bramley_et_al_2019, Verhavert_et_al_2019, Crompvoets_et_al_2022, vanDaal_et_al_2017, Gijsen_et_al_2021, Bouwer_et_al_2023},
none, to the best of the authors' knowledge, have provided a transparent
depiction of the DCJ system and the mechanisms generating the DCJ
outcome. This study aims to fill this gap by representing DCJ within the
framework of causal inference. Specifically, utilizing the structural
approach to causal inference, the study develops a scientific model to
clarify plausible causal assumptions and mechanisms inherent in the DCJ
system. The study then translates the scientific model into a
probabilistic statistical model. This model aims to produce statistical
estimates to draw inferences about plausible causal relationships within
the DCJ system.

@@

This section introduces fundamental concepts in causal inference, such
as directed acyclic graphs (DAGs), structural causal models (SCMs), and
the flow of association and causation in graphs. The section is not a
comprehensive description of causal inference methods. Readers
interested in deeper exploration should consult introductory papers like
\citet{Pearl_2010}, \citet{Rohrer_2018}, \citet{Pearl_2019}, and
\citet{Cinelli_et_al_2020}. They may also find introductory books such
as \citet{Pearl_et_al_2018}, \citet{Neal_2020} and
\citet{McElreath_2020} useful. For more advanced study, seminal
intermediate papers like \citet{Neyman_et_al_1923}, \citet{Rubin_1974},
\citet{Spirtes_et_al_1991}, and \citet{Sekhon_2009}, as well as books
like \citet{Pearl_2009}, \citet{Morgan_et_al_2014} and
\citet{Hernan_et_al_2020} are recommended.

Empirical research addresses real-world challenges by relying on
evidence gathered through observation and experimentation. In this
context, researchers typically frame their research questions as
\emph{estimands} or \emph{targets of inference}. These estimands
represent the specific quantities the study aims to determine
\citep{Everitt_et_al_2010}. For instance, a study might seek to answer
the question, ``To what extent do daily practice hours (\(X\)) influence
students' ability to produce high-quality written texts (\(Y\))?''. To
investigate this, the study randomly assigns students to two groups with
different levels of daily practice, low \((X=1)\) and medium \((X=2)\).
The quality of the written texts is evaluated using pairwise
comparisons, resulting in a dichotomous outcome \((Y=\{0,1\})\). The
research question is then framed as the estimand, ``\emph{On average},
which group of students produces higher-quality written texts?'' This
estimand is mathematically expressed as \(E[Y|X=1] - E[Y|X=2]\), where
\(E[\cdot]\) represents the expected value.

Researchers then proceed to identify the estimands.
\emph{Identification} is the process of accurately computing an estimand
using an estimator. An \emph{estimator} is a method or function that
transforms data into an estimate \citep{Neal_2020}. \emph{Estimates} are
numerical values that approximate the estimand and are derived through
\emph{estimation}, which is the process that integrates data with an
estimator \citep{Everitt_et_al_2010}. While various methods can
approximate an estimand, researchers prioritize estimators with
desirable properties that ensure accurate estimates. For example,
Fisher's exact test \citep{Fisher_1922} is an estimator with desirable
properties for comparing \(2\times2\) tables. The test provides accurate
estimates for comparing proportions when its assumptions are satisfied.
Figure~\ref{fig-IEflow}, known as the Identification-Estimation
flowchart \citep{McElreath_2020, Neal_2020}, illustrates the
aforementioned process of transitioning from estimands to estimates.

@@

\caption{\label{fig-IEflow}Identification-Estimation flowchart.
Extracted and slightly modified from \citet[32]{Neal_2020}}

@@

However, many studies aim to establish causal relationships rather than
merely infer associations. In the earlier example, the estimates
obtained using Fisher's exact test can be interpreted as causal because
the data were collected through a randomized experiment. In randomized
experiments, the researcher fully controls the treatment assignment
mechanism, such as assigning students to different groups. This control
allows causal estimates to be derived from associational estimates, such
as those provided by the Fisher statistic, since randomization
effectively eliminates confounding \citep{Neal_2020}. \emph{Confounding}
occurs when an external variable influences both the outcome and the
variable of interest, leading to spurious association
\citep{Everitt_et_al_2010}.

@@

\includegraphics[width=0.8\textwidth,height=\textheight]{images/figures/ACflow.png}

@@

\caption{\label{fig-ACflow}The flow of association and causation in
graphs. Extracted and slightly modified from \citet[31]{Neal_2020}}
